---
title: Optimal Portfolio Construction
author: Aaron Hardy
date: "`r Sys.Date()`"
slug: optimal-portfolio-construction
categories:
  - Stocks
tags:
  - Stocks
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
editor_options: 
  chunk_output_type: console
---
```{r include=FALSE}
knitr::opts_chunk$set(
      collapse=TRUE,
      comment="#>",
      fig.retina=2,
      message=FALSE,
      warning=FALSE)
```

# Optimal Portfolio Construction
 In this article we will compare several portfolio construction methodologies. We will evaluate the Modern Portfolio Theory and the Risk Parity portfolio and Resampled Efficient Frontier approaches.
 
The *Modern Portfolio Theory* approach attempts to maximize the return of a group of assets (a portfolio) for the level of risk borne by those assets. One complaint of this approach is its tendency to lead to overconcentrated portfolios.

The *Resampled Efficient Frontier* approach involves creating hundreds of efficient frontiers in an effort to create a more robust set of portfolio.

An alternative approach is the *Risk Parity Approach* which attempts to produce a group of assets (a portfolio) whose members contribute equally to the risk of the portfolio. This approach is hailed for its tendency to result in a more diverse group of portfolio assets than that of the Modern Portfolio Theory. 



```{r}

# Load packages
library(riskParityPortfolio)
library(fPortfolio)
library(tidyverse)
library(lubridate)
library(timetk)
library(tidyquant)
library(ggthemes) # For the economist theme
library(magrittr)
library(ggrepel)

# Define symbols to use
#   Stocks            ^GSPC represents the S&P 500 index
#   Bonds:            AGG represents investment-grade U.S. bonds
#   Commodities:      USCI represents diversified commodities such as oil, natural gas,                         gold, silver, aluminum, copper, corn, soybeans, live cattle,                              coffee, sugar, etc.
#   Real Estate:      VNQ represents U.S. REITs
#   Emerging Markets: VWO represents emerging markets
symbols <- c("^GSPC", "AGG", "USCI", "VNQ", "VWO")


# Get prices from Yahoo Finance
returns_monthly <- symbols %>%
  tq_get(get  = "stock.prices",
         from = "2004-12-31") %>%
  group_by(symbol) %>%
  tq_transmute(select     = adjusted,
            mutate_fun = periodReturn,
            period     = "monthly")

# Convert to xts object
returns_xts <- returns_monthly %>%
  pivot_wider(id_cols = date, names_from = symbol, values_from = monthly.returns) %>%
  drop_na() %>%
  tk_xts(select = -date, date_var = date)


# Calculate the covariance matrix
covariance <- cov(coredata(returns_xts))

# Compute the risk parity portfolio
portfolio_parity <- riskParityPortfolio(covariance)

# Compute the tangency portfolio
portfolio_tangency <- tangencyPortfolio(as.timeSeries(returns_xts),
                                        constraints = "LongOnly")
portfolio_weights <- rbind(portfolio_parity$w, getWeights(portfolio_tangency))
row.names(portfolio_weights) <- c("Parity Portfolio", "Tangency Portfolio")


portfolio_weights %>% 
  as.data.frame() %>%
  rownames_to_column("portfolio") %>%
  pivot_longer(-portfolio, names_to = "symbol", values_to = "weight") %>%
  ggplot(aes(x = symbol, y = weight)) + #, group =  portfolio)) +
  geom_bar(aes(fill = portfolio), stat = "identity", position = "dodge") +
  theme_economist()

returns_monthly %>% 
  mutate(cumulative.returns = cumprod(1 + monthly.returns)) %>%
  ggplot(aes(x = date, y = cumulative.returns)) + 
  geom_line(aes(color = symbol), size = 1.1, show.legend = FALSE) + 
  scale_x_date(#date_minor_breaks = "3 months",
               date_breaks= "1 year",
               date_labels = "%b %Y",
               expand = expand_scale(mult = c(0, .18))) +
  geom_label_repel(data = . %>% slice(n()),
                   aes(label = symbol, color = symbol),
                   nudge_x = 6,
                   min.segment.length = 0,
                   segment.color = NA,
                   direction = "x",
                   fontface = "bold", show.legend = FALSE) +
  labs(title = "Cumulative returns of various asset classes",
       y = "", x = "") +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.8, vjust = .5))   


tangency_wts <- portfolio_weights %>% extract("Tangency Portfolio", ) %>% round(2)
tangency_portfolio_returns <- returns_monthly %>%
  # drop_na() %>%
  # ungroup() %>%
  tq_portfolio(assets_col  = symbol,
               returns_col = monthly.returns, 
               weights     = tangency_wts, 
               col_rename  = "tangency_portfolio_return")


parity_wts <- portfolio_weights %>% extract("Parity Portfolio", ) %>% round(2)
parity_portfolio_returns <- returns_monthly %>%
  # drop_na() %>%
  # ungroup() %>%
  tq_portfolio(assets_col  = symbol,
               returns_col = monthly.returns, 
               weights     = parity_wts, 
               col_rename  = "parity_portfolio_return")


# Plot the cumulative returns of both strategies
parity_portfolio_returns %>%
  ggplot(aes(x = date, y = cumprod(1 + parity_portfolio_return))) +
  geom_line() +
  theme_economist()


```



